---
title: "Seurat colocalization locus plots"
output: html_notebook
---

This notebook follows the coloc data-structure vignette: each plot uses the
dense set of variants shared by the eQTL and GWAS datasets within one genomic
region. The publication plots show the eQTL association panel only, color
neighboring variants by 1000 Genomes European-reference LD with the requested
variant, highlight the requested variant in purple, and omit the gene track.

Reference: <https://cran.r-project.org/web/packages/coloc/vignettes/a02_data.html>

```{r setup}
suppressPackageStartupMessages({
  library(coloc)
  library(data.table)
  library(ggplot2)
  library(here)
})

here::i_am(".git/HEAD")

repo_root <- here()
target_root <- here(
  "processed-data", "11_eQTL_coloc", "seurat", "coloc",
  "susie_targeted_no23andMe", "all_targets_20260819"
)
target_summary_file <- file.path(
  target_root, "aggregate_eb_20260820", "baseline_vs_eb_target_summary.tsv.gz"
)
pvar_file <- file.path(target_root, "inputs", "target_pvar.tsv.gz")
plots_dir <- here("plots", "11_eQTL_coloc")

## primary manuscript examples requested by Jacqui.
plot_targets <- data.table(
  disorder = "BD",
  dataset_id = c("astro", "l2-3", "l5"),
  gene = "MAPK3",
  highlight = "rs55732507"
)

dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r helpers}
assert_files <- function(paths) {
  missing <- paths[!file.exists(paths)]
  if (length(missing) > 0L) {
    stop("Missing required file(s):\n", paste(missing, collapse = "\n"))
  }
  invisible(TRUE)
}

assert_cols <- function(dt, cols, label) {
  missing <- setdiff(cols, names(dt))
  if (length(missing) > 0L) {
    stop("Missing columns in ", label, ": ", paste(missing, collapse = ", "))
  }
  invisible(TRUE)
}

reference_gt_cache <- new.env(parent = emptyenv())

reference_gt_path <- function(chr) {
  file.path(
    target_root, "inputs", "reference",
    paste0(chr, ".1000G_highcov.EUR_unrelated.GT.tsv.gz")
  )
}

load_reference_gt <- function(chr) {
  if (exists(chr, envir = reference_gt_cache, inherits = FALSE)) {
    return(get(chr, envir = reference_gt_cache, inherits = FALSE))
  }
  path <- reference_gt_path(chr)
  assert_files(path)
  gt <- fread(path)
  assert_cols(gt, "variant_id", basename(path))
  assign(chr, gt, envir = reference_gt_cache)
  gt
}

reference_r2 <- function(variant_ids, index_variant_id, chr) {
  gt <- load_reference_gt(chr)
  keep <- gt[variant_id %chin% unique(c(variant_ids, index_variant_id))]
  if (!index_variant_id %chin% keep$variant_id) {
    stop("LD index variant is absent from the reference genotypes: ", index_variant_id)
  }

  sample_cols <- setdiff(names(keep), "variant_id")
  genotype <- as.matrix(keep[, ..sample_cols])
  storage.mode(genotype) <- "numeric"
  index_genotype <- genotype[match(index_variant_id, keep$variant_id), ]
  r <- as.numeric(stats::cor(t(genotype), index_genotype, use = "pairwise.complete.obs"))
  data.table(
    variant_id = keep$variant_id,
    r2 = pmin(1, pmax(0, r^2)),
    ld_reference_n = length(sample_cols)
  )
}

local_nominal_path <- function(path) {
  if (file.exists(path)) return(path)
  file.path(target_root, "inputs", "nominal", basename(path))
}

resolve_coloc_target <- function(gene, disorder, dataset_id = NULL,
                                 context = NULL, target_id = NULL) {
  gene_value <- gene
  disorder_value <- disorder
  dataset_id_value <- dataset_id
  context_value <- context
  target_id_value <- target_id
  rows <- copy(target_summary)
  if (!is.null(target_id_value)) rows <- rows[target_id == target_id_value]
  rows <- rows[
    disorder == disorder_value &
      (gene_id == gene_value | gene_name == gene_value)
  ]
  if (!is.null(dataset_id_value)) rows <- rows[dataset_id == dataset_id_value]
  if (!is.null(context_value)) rows <- rows[context == context_value]

  if (nrow(rows) != 1L) {
    stop(
      "Expected one coloc target for gene=", gene,
      ", disorder=", disorder,
      if (!is.null(dataset_id)) paste0(", dataset_id=", dataset_id) else "",
      if (!is.null(context)) paste0(", context=", context) else "",
      "; found ", nrow(rows), "."
    )
  }
  rows[1]
}

resolve_highlights <- function(x, highlight) {
  if (is.null(highlight) || length(highlight) == 0L) return(character())
  highlight <- unique(as.character(highlight))
  variant_hits <- highlight[highlight %chin% x$plot_data$variant_id]
  rsid_hits <- x$plot_data[rsid %chin% highlight, unique(variant_id)]
  resolved <- unique(c(variant_hits, rsid_hits))
  missing <- setdiff(highlight, c(variant_hits, x$plot_data[variant_id %chin% resolved, rsid]))
  if (length(missing) > 0L) {
    stop("Highlight variant(s) not found in the coloc region: ", paste(missing, collapse = ", "))
  }
  resolved
}

load_coloc_target <- function(gene, disorder, dataset_id = NULL,
                              context = NULL, target_id = NULL,
                              p1 = 1e-4, p2 = 1e-4, p12 = 1e-5) {
  target <- resolve_coloc_target(
    gene = gene, disorder = disorder, dataset_id = dataset_id,
    context = context, target_id = target_id
  )
  nominal_file <- local_nominal_path(target$nominal_file)
  gwas_file <- file.path(target_root, "inputs", "gwas", paste0(disorder, "_approved.tsv.gz"))
  assert_files(c(nominal_file, gwas_file, pvar_file))

  eqtl <- fread(nominal_file)
  gwas <- fread(gwas_file)
  pvar <- fread(pvar_file)
  assert_cols(
    eqtl,
    c("phenotype_id", "variant_id", "af", "pval_nominal", "slope", "slope_se"),
    basename(nominal_file)
  )
  assert_cols(
    gwas,
    c("variant_id", "rsid", "beta", "beta_se", "N", "ncas", "p"),
    basename(gwas_file)
  )
  assert_cols(pvar, c("variant_id", "CHROM", "POS"), basename(pvar_file))

  ## coloc uses only variants available in both studies within one dense region.
  plot_data <- merge(eqtl, gwas, by = "variant_id", suffixes = c("_eqtl", "_gwas"))
  plot_data <- merge(
    plot_data,
    pvar[, .(variant_id, chromosome = CHROM, position = POS)],
    by = "variant_id"
  )
  plot_data <- plot_data[
    position >= target$region_start & position <= target$region_end &
      is.finite(pval_nominal) & pval_nominal > 0 & pval_nominal <= 1 &
      is.finite(p) & p > 0 & p <= 1
  ]
  setorder(plot_data, position, variant_id)
  if (nrow(plot_data) < 10L) stop("Too few shared variants for coloc plotting.")
  if (anyDuplicated(plot_data$variant_id)) stop("Duplicate shared variant IDs in coloc plotting data.")

  dataset1 <- list(
    beta = plot_data$slope,
    varbeta = plot_data$slope_se^2,
    pvalues = plot_data$pval_nominal,
    snp = plot_data$variant_id,
    position = plot_data$position,
    N = target$n_eqtl,
    MAF = pmin(plot_data$af, 1 - plot_data$af),
    type = "quant"
  )
  dataset2 <- list(
    beta = plot_data$beta,
    varbeta = plot_data$beta_se^2,
    pvalues = plot_data$p,
    snp = plot_data$variant_id,
    position = plot_data$position,
    N = stats::median(plot_data$N, na.rm = TRUE),
    type = "cc",
    s = stats::median(plot_data$ncas / plot_data$N, na.rm = TRUE)
  )
  coloc::check_dataset(dataset1)
  coloc::check_dataset(dataset2)

  fit <- NULL
  invisible(utils::capture.output(
    fit <- coloc::coloc.abf(
      dataset1 = dataset1, dataset2 = dataset2,
      p1 = p1, p2 = p2, p12 = p12
    )
  ))
  plot_data[fit$results, SNP.PP.H4 := i.SNP.PP.H4, on = .(variant_id = snp)]
  plot_data[, `:=`(
    neglog10_eqtl = -log10(pval_nominal),
    neglog10_gwas = -log10(p)
  )]

  ## retain both the fresh result and the manuscript target metadata.
  list(
    target = target,
    dataset1 = dataset1,
    dataset2 = dataset2,
    fit = fit,
    plot_data = plot_data
  )
}

r2_legend_grob <- function() {
  colors <- c("#000080", "#2C7FB8", "#32CD32", "#FFA500", "#FF0000")
  bar_left <- 0.15
  bar_right <- 0.43
  edges <- seq(0.08, 0.75, length.out = length(colors) + 1L)
  children <- list(
    grid::rectGrob(
      x = 0.5, y = 0.5, width = 0.98, height = 0.98,
      gp = grid::gpar(fill = "white", col = "grey50", lwd = 0.8)
    ),
    grid::textGrob(
      expression(r^2), x = 0.5, y = 0.88,
      gp = grid::gpar(col = "grey15", fontsize = 10)
    )
  )

  ## draw bordered blocks so internal boundaries span the full cassette width.
  for (i in seq_along(colors)) {
    children[[length(children) + 1L]] <- grid::rectGrob(
      x = (bar_left + bar_right) / 2,
      y = (edges[i] + edges[i + 1L]) / 2,
      width = bar_right - bar_left,
      height = edges[i + 1L] - edges[i],
      gp = grid::gpar(fill = colors[i], col = "grey20", lwd = 0.7)
    )
  }

  ## ticks sit outside the cassette on the label side only.
  tick_y <- edges[2:5]
  children[[length(children) + 1L]] <- grid::segmentsGrob(
    x0 = bar_right, x1 = 0.51, y0 = tick_y, y1 = tick_y,
    gp = grid::gpar(col = "grey20", lwd = 0.7)
  )
  children[[length(children) + 1L]] <- grid::textGrob(
    c("0.2", "0.4", "0.6", "0.8"),
    x = 0.58, y = tick_y, just = "left",
    gp = grid::gpar(col = "grey15", fontsize = 8.5)
  )
  grid::gTree(children = do.call(grid::gList, children))
}

plot_coloc_locus <- function(x, highlight = NULL,
                             trait = c("eQTL", "GWAS", "both"),
                             color_by = c("r2", "highlight"),
                             point_color = "#17365D",
                             highlight_color = "#7B2CBF",
                             point_size = 2.2,
                             title = "auto",
                             subtitle = "auto") {
  trait <- match.arg(trait)
  color_by <- match.arg(color_by)
  highlight_ids <- resolve_highlights(x, highlight)
  if (color_by == "r2" && length(highlight_ids) != 1L) {
    stop("r2 coloring requires exactly one highlighted index variant.")
  }
  dt <- copy(x$plot_data)
  dt[, highlighted := variant_id %chin% highlight_ids]
  if (color_by == "r2") {
    ld <- reference_r2(
      variant_ids = dt$variant_id,
      index_variant_id = highlight_ids,
      chr = x$target$chr
    )
    dt[ld, `:=`(r2 = i.r2, ld_reference_n = i.ld_reference_n), on = "variant_id"]
  } else {
    dt[, `:=`(r2 = NA_real_, ld_reference_n = NA_integer_)]
  }

  long <- if (trait == "eQTL") {
    dt[, .(variant_id, rsid, position, highlighted, r2, ld_reference_n,
           trait = "eQTL", value = neglog10_eqtl)]
  } else if (trait == "GWAS") {
    dt[, .(variant_id, rsid, position, highlighted, r2, ld_reference_n,
           trait = "GWAS", value = neglog10_gwas)]
  } else {
    rbindlist(list(
      dt[, .(variant_id, rsid, position, highlighted, r2, ld_reference_n,
             trait = "eQTL", value = neglog10_eqtl)],
      dt[, .(variant_id, rsid, position, highlighted, r2, ld_reference_n,
             trait = "GWAS", value = neglog10_gwas)]
    ))
    long[, trait := factor(trait, levels = c("eQTL", "GWAS"))]
  }

  target <- x$target
  pp4 <- as.numeric(x$fit$summary[["PP.H4.abf"]])
  auto_title <- paste0(
    target$gene_name, " eQTL-", target$disorder, " GWAS colocalization"
  )
  title_to_use <- if (identical(title, "auto")) auto_title else title
  highlight_label <- if (length(highlight_ids) == 1L) {
    row <- dt[variant_id == highlight_ids][1]
    if (!is.na(row$rsid) && nzchar(row$rsid)) row$rsid else row$variant_id
  } else if (length(highlight_ids) > 1L) {
    paste(length(highlight_ids), "highlighted variants")
  } else {
    "no highlighted variant"
  }
  auto_subtitle <- sprintf(
    "%s | highlighted variant: %s | coloc PP4 = %.3f",
    target$context, highlight_label, pp4
  )
  subtitle_to_use <- if (identical(subtitle, "auto")) auto_subtitle else subtitle

  p <- ggplot(long, aes(x = position / 1e6, y = value))
  if (color_by == "r2") {
    p <- p +
      geom_point(
        data = long[highlighted == FALSE],
        aes(fill = r2), color = "grey20", shape = 21,
        size = point_size, stroke = 0.3, alpha = 0.88
      ) +
      scale_fill_stepsn(
        ## hold each LD bin at one color, matching the reference locus plot.
        colours = rep(c("#000080", "#2C7FB8", "#32CD32", "#FFA500", "#FF0000"), each = 2L),
        values = c(0, 0.1999, 0.2, 0.3999, 0.4, 0.5999, 0.6, 0.7999, 0.8, 1),
        breaks = c(0.2, 0.4, 0.6, 0.8),
        limits = c(0, 1),
        na.value = "grey75",
        guide = "none"
      )
  } else {
    p <- p + geom_point(
      data = long[highlighted == FALSE],
      fill = point_color, color = "grey20", shape = 21,
      size = point_size, stroke = 0.3, alpha = 0.82
    )
  }
  p <- p +
    geom_point(
      data = long[highlighted == TRUE],
      color = "grey20", fill = highlight_color,
      size = point_size + 0.3, shape = 23, stroke = 0.4
    ) +
    labs(
      title = title_to_use,
      subtitle = subtitle_to_use,
      x = paste0("Chromosome ", sub("^chr", "", target$chr), " position (Mb)"),
      y = expression(-log[10](italic(P)))
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.12))) +
    theme_classic(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(size = 10),
      strip.text.y = element_text(face = "bold"),
      axis.line = element_line(linewidth = 0.4),
      legend.position = "none"
    )
  if (trait == "both") p <- p + facet_grid(trait ~ ., scales = "free_y")

  label_data <- long[highlighted == TRUE]
  if (nrow(label_data) > 0L) {
    label_data[, label := ifelse(!is.na(rsid) & nzchar(rsid), rsid, variant_id)]
    p <- p + geom_text(
      data = label_data,
      aes(label = label),
      color = "grey15", nudge_y = 0.35,
      vjust = 0, size = 3.2, fontface = "bold",
      check_overlap = TRUE
    )
  }
  if (color_by == "r2") {
    x_range <- range(long$position / 1e6, na.rm = TRUE)
    y_range <- range(c(0, long$value), na.rm = TRUE)
    x_span <- diff(x_range)
    y_span <- diff(y_range)
    legend_xmax <- x_range[2] - 0.01 * x_span
    legend_ymax <- y_range[2] + 0.105 * y_span
    p <- p + annotation_custom(
      r2_legend_grob(),
      xmin = legend_xmax - 0.14 * x_span,
      xmax = legend_xmax,
      ymin = legend_ymax - 0.45 * y_span,
      ymax = legend_ymax
    )
  }

  list(
    plot = p,
    data = long,
    highlight_ids = highlight_ids,
    ld_source = if (color_by == "r2") {
      paste0("1000 Genomes high-coverage EUR unrelated; n=", unique(na.omit(long$ld_reference_n)))
    } else {
      NA_character_
    }
  )
}

save_coloc_pdf_set <- function(plots, filename, width = 6.5, height = 4.2) {
  out_file <- file.path(plots_dir, filename)
  grDevices::pdf(out_file, width = width, height = height, onefile = TRUE)
  on.exit(grDevices::dev.off(), add = TRUE)
  for (plot in plots) print(plot)
  out_file
}
```

```{r load-inputs}
assert_files(c(target_summary_file, pvar_file))
target_summary <- fread(target_summary_file)
assert_cols(
  target_summary,
  c(
    "target_id", "dataset_id", "gene_id", "gene_name", "disorder", "context",
    "chr", "nominal_file", "region_start", "region_end", "n_eqtl", "PP4"
  ),
  basename(target_summary_file)
)
```

```{r mapk3-bd-plots, fig.width=6.5, fig.height=4.2}
mapk3_results <- vector("list", nrow(plot_targets))

for (i in seq_len(nrow(plot_targets))) {
  spec <- plot_targets[i]
  coloc_target <- load_coloc_target(
    gene = spec$gene,
    disorder = spec$disorder,
    dataset_id = spec$dataset_id
  )
  pobj <- plot_coloc_locus(
    coloc_target,
    highlight = spec$highlight,
    trait = "eQTL"
  )

  coloc_target$plot_result <- pobj
  mapk3_results[[i]] <- coloc_target
  print(pobj$plot)
}
names(mapk3_results) <- plot_targets$dataset_id

mapk3_pdf <- save_coloc_pdf_set(
  lapply(mapk3_results, function(x) x$plot_result$plot),
  "coloc_MAPK3_BD_rs55732507_Astro_L23_L5.pdf"
)
```

```{r write-mapk3-summary}
mapk3_summary <- rbindlist(lapply(mapk3_results, function(x) {
  highlight_id <- x$plot_result$highlight_ids[[1]]
  highlight_row <- unique(x$plot_result$data[variant_id == highlight_id])
  data.table(
    disorder = x$target$disorder,
    dataset_id = x$target$dataset_id,
    context = x$target$context,
    gene_id = x$target$gene_id,
    gene_name = x$target$gene_name,
    highlight_variant_id = highlight_id,
    highlight_rsid = highlight_row$rsid[[1]],
    highlight_r2 = highlight_row$r2[[1]],
    ld_source = x$plot_result$ld_source,
    nsnps = as.integer(x$fit$summary[["nsnps"]]),
    PP4 = as.numeric(x$fit$summary[["PP.H4.abf"]])
  )
}))
mapk3_summary[, plot_rank := match(dataset_id, plot_targets$dataset_id)]
setorder(mapk3_summary, plot_rank)
mapk3_summary[, plot_rank := NULL]

mapk3_summary
```

```{r validate-outputs}
expected_pp4 <- setNames(
  target_summary[
    disorder == "BD" & gene_name == "MAPK3" & dataset_id %chin% plot_targets$dataset_id,
    PP4
  ],
  target_summary[
    disorder == "BD" & gene_name == "MAPK3" & dataset_id %chin% plot_targets$dataset_id,
    dataset_id
  ]
)
observed_pp4 <- setNames(mapk3_summary$PP4, mapk3_summary$dataset_id)
stopifnot(
  file.exists(mapk3_pdf),
  identical(mapk3_summary$dataset_id, plot_targets$dataset_id),
  all(mapk3_summary$nsnps == 1561L),
  all(mapk3_summary$highlight_r2 == 1),
  all(grepl("1000 Genomes.*n=503", mapk3_summary$ld_source)),
  isTRUE(all.equal(observed_pp4[names(expected_pp4)], expected_pp4, tolerance = 1e-10))
)
```

```{r generic-examples, eval=FALSE}
## plot another targeted gene/context/disorder combination.
example_target <- load_coloc_target(
  gene = "FBLN7",
  disorder = "MDD",
  dataset_id = "astro"
)
plot_coloc_locus(example_target, highlight = example_target$target$lead_rsid)$plot

## show aligned eQTL and GWAS association panels when both are useful.
plot_coloc_locus(example_target, highlight = example_target$target$lead_snp, trait = "both")$plot
```
